---
output: 
    md_document:
        variant: markdown_strict+fenced_code_blocks
        includes:
            in_header: hdr
title: Dot-density maps
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      message = FALSE)
knitr::knit_hooks$set(source = function(x, options) {
    x <- paste0("\n~~~~~r\n", paste0(x, collapse = "\n"), "\n~~~~~\n")
    x
})
```
# Dot Density maps

A [dot-density map](https://en.wikipedia.org/wiki/Dot_distribution_map) is one way to map aggregated spatial data without some of the distortions inherent in choropleths. Several recent tools in R, in particular the `tidycensus` (for demographic data), `tigris` (for spatial shape files), and `sf` (for manipulating geospatial data) pacakges. What follows is a short tutorial on creating a dot-density map, using as an example the distribution of educational attainment in Alameda County, CA. 

## Acquiring the data

I'll use the `tidycensus::get_acs` function to pull the data at the appropriate geographic level, but first I need to figure out the name of the table that contains the educational attainment numbers:

```{r}
# load libraries
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
options(tigris_use_cache=TRUE)
options(tigris_class="sf")
```

```{r, cache = TRUE}
v16 <- load_variables(2016, "acs5", cache = TRUE)
v16 %>%
    mutate(table = str_extract(name, "^.+_")) %>%
    filter(str_detect(concept, "EDUCATIONAL ATTAINMENT")) %>%
    select(table, concept) %>% distinct %>% print(n = Inf)
```

I'll be using table `B15003`: "EDUCATIONAL ATTAINMENT FOR THE POPULATION 25 YEARS AND OVER". Note that that means that my map will be ignoring the existence of children.

Next I pull the data. By including `geometry = TRUE` in the `get_acs` call, I automatically get an `sf` object that includes the associated geometry along with the demgoraphic data. The state/county codes are FIPS codes:

```{r, cache = TRUE}
acs <- get_acs("tract", table = "B15003", cache_table = TRUE,
               geometry = TRUE, state = "06", county = "001",
               year = 2016, output = "tidy")
acs
```

The educational attainment splits things out to quite a few levels (with one for "finished 4th grade" and another for "finished 5th grade" and so on), so I'll collapse them down to a handful of categories.

```{r, cache = TRUE}
acs <- acs %>%
    mutate(
        id = str_extract(variable, "[0-9]{3}$") %>% as.integer
    ) %>%
    mutate(education =case_when(
        id %>% between(2, 16) ~ "No HS diploma",
        id %>% between(17, 21) ~ "HS, no Bachelors",
        id == 22 ~ "Bachelors",
        id > 22 ~ "Post-Bachelors"
    )) %>% filter(!is.na(education)) %>%
    group_by(GEOID, education) %>% 
    summarise(estimate = sum(estimate))
acs
```

Dot-density maps work by placing dots randomly within the appropriate geographic boundaries, to approximate the overall distribution of people in space. The function `sf::st_sample` samples points from within polygons. It seems like there are different ways of approaching the next step, what I did was split the data by education level, and then run the sampling function on each level for each block group. Grouping and summarizing at the end gets me to 4 multipoint layers, each containing a number of points. I also re-code the education level to an ordered factor variable, which will help when plotting:

```{r, cache = TRUE}
acs_split <- acs %>%
    filter(estimate > 50) %>%
    split(.$education)

generate_samples <- function(data) suppressMessages(
    data %>% st_sample(size = round(.$estimate / 100)))

points <- map(acs_split, generate_samples)
points <- imap(points, ~st_sf(data_frame(education = rep(.y, length(.x))),
                         geometry = .x))
points <- do.call(rbind, points)
points <- points %>% group_by(education) %>% summarise()
points <- points %>%
    mutate(education = factor(
        education,
        levels = c("No HS diploma", "HS, no Bachelors",
                   "Bachelors", "Post-Bachelors")))
points %>% mutate(n_points = map_int(geometry, length))
```

## Plotting
```{r}
# setting theme options
theme_set(theme_minimal() +
              theme(panel.grid.major = element_line(size = .1),
                    axis.title = element_blank(),
                    text = element_text(family = "Roboto Condensed"),
                    panel.grid.major.x = element_line(size = 0),
                    panel.grid.major.y = element_line(size = 0),
                    axis.text = element_blank(),
                    legend.position = "bottom"))
```
```{r}
ggplot() + 
    geom_sf(data = points, 
            aes(colour = education,
                fill = education),
            size = .1) + 
    scale_color_brewer(type = "div", palette = 4) + 
    scale_fill_brewer(type = "div", palette = 4)
```

That's a pretty good start. The `tigris` package makes it easy to get various geography layers from TIGER, I'll add water, major roads, and label the towns in the county. I'll also pull down the outline of Alameda County:
```{r, cache = TRUE}
water <- tigris::area_water("06", "001")
towns <- tigris::county_subdivisions("06", county = "001")
alameda_roads <- tigris::roads("06", "001")
ca_county <- tigris::counties(state = "06")
alameda <- ca_county %>% filter(COUNTYFP == "001")

# create town labels by finding the centroid of each town
# ggplot's label functions work better with X/Y dataframes rather 
# than sf objects
town_labels <- towns %>% select(NAME) %>%
    mutate(center = st_centroid(geometry)) %>%
    as.tibble %>%
    mutate(center = map(center, ~st_coordinates(.) %>%
                            as_data_frame)) %>%
    select(NAME, center) %>% unnest()
```

Plotting with ggplot, using `geom_sf`:
```{r, cache = TRUE, fig.width = 11, fig.height = 8, fig.align="center", class.output="full-width", dpi = 200}
ggplot() +
    geom_sf(data = alameda, size = .1, fill = NA) +
    geom_sf(data = water, colour = "#eef7fa", size = .1,
            fill = "#e6f3f7") +
    geom_sf(data = points,
            aes(colour = education, fill = education),
            size = .1) +
    geom_sf(data = alameda_roads %>% filter(RTTYP %in% c("I", "S")),
            size = .2, colour = "gray40") +
    scale_color_brewer(type = "div", palette = 4) +
    scale_fill_brewer(type = "div", palette = 4) +
    ggrepel::geom_label_repel(data = town_labels,
                             aes(x = X, y = Y, label = NAME),
                             size = 3, family = "Roboto Condensed",
                             label.padding = unit(.1, "lines"), alpha = .7)
```

