---
output: 
    md_document:
        includes:
            in_header: hdr
title: Using functions to implement a stack
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Stacks
A stack is a data structure with just a couple of operations: 

- push: push a new item onto the top of the stack
- pop: return the most recently pushed item, while removing it from the stack as a side-effect.

In a purely functional setting, we'd like to avoid side-effects. So in this document I'll be following the convention of [this implementation](https://github.com/oneilsh/rstackdeque) and use the following operations: 

- push: return a new stack that is the result of pushing a new item onto a stack, while leaving the original stack unchanged
- peak: return the most recently pushed item on the stack, and leave the stack unchanged
- without_top: return a new stack that is the result of removing the most recently pushed item from the top of a stack, while leaving the original stack unchanged
- size: return the number of elements in a stack

### Functions

I linked above to [an implementation](https://github.com/oneilsh/rstackdeque) by Shawn Oneil that implements a persistent, functional stack by explicitly using environments. This implementation is basically the same, but by relying just on function calls, I'm hoping it will reveal a bit more of the way functions work. 

## Set up

First we set up the API for the stack data structure:

```{r api}
peak <- function(s) UseMethod("peak")
push <- function(s, new_item) UseMethod("push")
without_top <- function(s) UseMethod("without_top")
size <- function(s) UseMethod("size")
```

And it would probably be helpful to have a reasonable print method

```{r print-stack}
print.stack <- function(s) {
    cat("A stack of size", size(s))
    if (size(s) > 0L) {
        cat("\nTop:\n") 
        str(peak(s))
    }
    invisible(s)
}
```

## The base case

And now we implement a constructor for an empty stack. In the process, we end up implementing `peak`, `without_top`, and `size`. 

```{r empty-stack}
# an empty stack should just be a function that returns the expected values
stack <- function() {
    f <- function(query) 
        switch(query, 
               peak = NULL,
               without_top = stop("Stack is already empty", call. = FALSE),
               size = 0L)
    structure(f, class = "stack")
}

peak.stack <- function(s) s("peak")
without_top.stack <- function(s) s("without_top")
size.stack <- function(s) s("size")
```

All that is left is to implement `push`.

## Push

The `push` operation should return a new stack. Note that `without_top` on the newly created stack should just return the stack that was used in the call to `push`, so we really don't need to do much. The one consideration to make here is to ensure that `size(s)` gets evaluated. Otherwise, every call to get the `size` of a stack would result in recursively calling for the size of every stack below it. :

```{r push}
push.stack <- function(s, new_item) {
    sz <- size(s) + 1L
    f <- function(query) 
        switch(query,
               peak = new_item,
               without_top = s,
               size = sz)
    structure(f, class = "stack")
}
```

That's it! We have implemented a stack. There's nothing here that's super sophisticated, but the syntax feels a little magical. In [Oneil's implementation](https://github.com/oneilsh/rstackdeque) he uses environments to assign values, and here I rely on the fact that [functions carry around their own environments](http://adv-r.had.co.nz/Functions.html#function-components). We still share data to achieve speed and efficiency, along with persistence.

## Test it out

Let's try it out:

```{r stack-tests}
# an empty stack
s <- stack()
size(s)
t <- push(s, 1)
u <- push(t, 2)
peak(t)
peak(u)

# shared data
identical(without_top(u), t)

# as opposed to:
t2 <- push(s, 1)
identical(t2, t)
```

## Convenience methods

Making an `as.list` method makes the structure `lapply`-able, among other things. 

```{r as-list}
as.list.stack <- function(s) {
    # pre-allocate
    res <- vector("list", size(s))
    
    for (ind in seq_len(size(s))) {
        res[[ind]] <- peak(s)
        s <- without_top(s)
    }
    res
}
```

